---
title: "Baseball data: Learning Activity 4"
format: pdf
editor: visual
---

## Background

This activity uses the **baseball** dataset (available in the *SDAResources* package). The dataset contains statistics on **797 baseball players** from the rosters of all major league teams in November 2004 (Forman, 2004).

You will treat this file as a **population** and draw samples from it using stratified random sampling.

## Getting Started

Loading the dataset and the R packages. Visit the solutions for learning activity 3 on R introductions.

```{r,message=FALSE}
#| label: setup
# Uncomment the install.packages("SDAResources") line to install the package. 
# Comment it back after you install the package :) 
# install.packages("SDAResources") 
library(SDAResources)
library(dplyr)

data(baseball)         

```

## Exercise 1: Stratified random sample

Take a stratified random sample of 150 players from the population, using proportional allocation with the different teams as strata. Save your sample as `baseball_strs` for use in future exercises. Don't forget to set your seed. I've included some code to get you started. NOTE this chunk won't run until you fill in the missing blanks.

```{r}
# Set seed for reproducibility
set.seed(123)

# Total sample size
n <- 150

# Total population size
N <- nrow(baseball)

# Step 1: Get stratum sizes (Nh = number of players per team)
prop_alloc <- baseball %>% 
  group_by(team) %>% 
  summarise(Nh = n())

# Step 2: Calculate proportional sample size for each stratum
prop_alloc <- prop_alloc %>%
  mutate(nh = round(____________))   

# Step 3: Create an empty data frame to store the sample
baseball_strs <- data.frame()

# Step 4: Loop through each team and sample nh players
for (i in 1:nrow(prop_alloc)) {
  
  # Get the current team name
  current_team <- prop_alloc$team[i]          
  
  # Get the sample size for this team
  current_nh <- prop_alloc$nh[i]   
  
  # Get the Stratum size for this team
  current_Nh <- prop_alloc$Nh[i]   
  
  # Filter the data to only include the current team
  team_data <- baseball %>%
    filter(team == current_team)                  
  
  # Take a random sample of current_nh players from this team
  team_sample <- team_data[sample(1:nrow(team_data), size = ____________), ]   
  
  # Append the stratum size and sample size so you have easy access for calculations later
  team_sample$nh = current_nh
  team_sample$Nh = current_Nh
  
  # Add the team sample to our growing stratified sample
  baseball_strs <- rbind(baseball_strs, team_sample) 
}


```

```{r}
head(baseball_strs)
```

Describe how you selected the sample here:

## Exercise 2: Inference for the population proportion

Estimate the population proportion of players who are pitchers and give a 95% confidence interval, using your stratified sample. Do this manually using the formulas from lecture/the textbook — do not rely on pre-built R functions except for `mean()`, `sd()`, `qnorm(0.975)`, or `qt(0.975, df = n - 1)`, which give you the 97.5th quantile from a standard normal distribution and a t-distribution with $n - 1$ degrees of freedom, respectively.

```{r}
## Include your code here

df_forcalc <- baseball_strs %>% 
  group_by(team) %>% 
  summarise(phat = mean(pos == "P"),
            nh = nh[1], # just grabs the first one (they are all the same for a specific team)
            Nh = Nh[1]) # just grabs the first one (they are all the same for a specific team)


### Include your code here to calculate the confidence interval 

phat_str = ____________
variance_hat = ____________

ci_lwr = ____________
ci_upr = ______________

ci_lwr
ci_upr

```

Give the estimate and the 95% CI and interpret them here:

## Exercise 3: Inference for the population mean

Estimate the population mean of `logsal` and give a 95% CI, using your stratified sample with proportional allocation. Recall from learning activity 3 that `logsal` is the logarithm of `salary`. Assume the strata population variances are unknown here.

Do this manually using the formulas from lecture/the textbook — do not rely on pre-built R functions except for `mean()`, `sd()`, `qnorm(0.975)`, or `qt(0.975, df = n - 1)`, which give you the 97.5th quantile from a standard normal distribution and a t-distribution with $n - 1$ degrees of freedom, respectively.

```{r}
# Include your code here

baseball_strs <- baseball_strs %>% 
  mutate(logsal = log(salary))

df_forcalc <- baseball_strs %>% 
  group_by(team) %>% 
  summarise(ymean = mean(logsal),
            sd = sd(logsal),
            nh = nh[1], # just grabs the first one (they are all the same for a specific team)
            Nh = Nh[1]) # just grabs the first one (they are all the same for a specific team)

df_forcalc %>% head(5)


### Include your code here to calculate the confidence interval 

ymean_str = ____________________
variance_hat = ________________________

ci_lwr = ___________________
ci_upr = ____________________

ci_lwr
ci_upr

```

Give the estimate and the 95% CI and interpret them here:

## Exercise 4: Comparison to CI from Learning Activity 3

How do your estimates compare with those of Learning activity 3?

## Exercise 5: Justification for optimal allocation

Examine the population standard deviation in each stratum. I've included some code for you to fill out.

```{r}
## Use this for code

sh <- baseball %>% 
  mutate(logsal = log(salary)) %>% 
  group_by(team) %>% 
  summarise(sd = sd(logsal))

sh


```

Do you think optimal allocation would be worthwhile for this problem? Justify your response.

## Exercise 6: Computing optimal allocations

Use the population standard deviations from Exercise 5 to determine the optimal allocation for a sample of size 150.

```{r}

opt_alloc <- prop_alloc  

# Include your code here

# Adding the stratas population standard deviations
opt_alloc <- opt_alloc %>% 
  left_join(sh, by = "team")

# Calculate optimal allocation sample size for each stratum
opt_alloc <- opt_alloc %>%
  mutate(nh_opt = round( _________________))   

head(opt_alloc)
```

How much does the optimal allocation differ from the proportional allocation? Do you observe any patterns in how observations were re-allocated?

## Exercise 7: Stratified random sample using optimal allocation

Take a stratified random sample of 150 players from the population, using optimal allocation with the different teams as strata. Save your sample as `baseball_strs_opt` for use in future exercises. Don't forget to set your seed. I've included some code to get you started. NOTE this chunk won't run until you fill in the missing blanks.

```{r}
# Set seed for reproducibility
set.seed(123)

# Step 3: Create an empty data frame to store the sample
baseball_strs_opt <- data.frame()

# Step 4: Loop through each team and sample nh players
for (i in 1:nrow(opt_alloc)) {
  
  # Get the current team name
  current_team <- opt_alloc$team[i]          
  
  # Get the sample size for this team
  current_nh <- opt_alloc$nh_opt[i]   
  
  # Get the Stratum size for this team
  current_Nh <- opt_alloc$Nh[i]   
  
  # Filter the data to only include the current team
  team_data <- baseball %>%
    filter(team == current_team)               
  
  # Take a random sample of current_nh players from this team
  team_sample <- team_data[sample(1:nrow(team_data), size = _____________), ]   
  
  # Append the stratum size and sample size so you have easy access for calculations later
  team_sample$nh = current_nh
  team_sample$Nh = current_Nh
  
  # Add the team sample to our growing stratified sample
  baseball_strs_opt <- rbind(baseball_strs_opt, team_sample) 
}


```

```{r}
head(baseball_strs_opt)
```

## Exercise 8: Inference for the population mean under optimal allocation

Estimate the population mean of `logsal` and give a 95% CI, using your stratified sample with optimal allocation. Assume strata population variances are unknown.

Do this manually using the formulas from lecture/the textbook — do not rely on pre-built R functions except for `mean()`, `sd()`, `qnorm(0.975)`, or `qt(0.975, df = n - 1)`, which give you the 97.5th quantile from a standard normal distribution and a t-distribution with $n - 1$ degrees of freedom, respectively.

```{r}
# Include your code here

baseball_strs_opt <- baseball_strs_opt %>% 
  mutate(logsal = log(salary))

df_forcalc <- baseball_strs_opt %>% 
  group_by(team) %>% 
  summarise(ymean = mean(logsal),
            sd = sd(logsal),
            nh = nh[1], # just grabs the first one (they are all the same for a specific team)
            Nh = Nh[1]) # just grabs the first one (they are all the same for a specific team)

df_forcalc %>% head(5)


### Include your code here to calculate the confidence interval 

ymean_str = ____________________
variance_hat = ____________________

ci_lwr = ____________________
ci_upr = ____________________

ci_lwr
ci_upr





```

Give the estimate and the 95% CI and interpret them here. How do they compare with the confidence interval obtained in exercise 3?
