---
title: "Learning activity 5: Ratio Estimators with Real Census Data"
subtitle: "Using IPUMS ACS 2024 Data for New York State"
format: pdf
editor: visual
---

## Introduction

In class, we studied the **ratio estimator** under simple random sampling (SRS). In this activity, you will apply that theory to real survey data from the American Community Survey (ACS), accessed through [IPUMS USA](https://usa.ipums.org).

The key idea behind the ratio estimator is that if we know the population total of some auxiliary variable $x$, and $x$ is correlated with our variable of interest $y$, we can use that relationship to get a better estimate of the population total of $y$.

The ratio estimator for the population total is:

$$\hat{\bar{y}}_r = \hat{B} \cdot t_x \quad \text{where} \quad \hat{B} = \frac{\bar{y}_S}{\bar{x}_S}$$

and $t_x$ is the **known** population total of $x$. In this activity we use **annual hours worked** as our auxiliary variable and **wage income** as our variable of interest. We treat the full filtered ACS dataset as the population, draw an SRS from it, and then verify our estimates against the true population values.

------------------------------------------------------------------------

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(ipumsr)
library(survey)
```

```{r, message=FALSE, warning=FALSE}
load('datACS.RData')
```

------------------------------------------------------------------------

## Prepare the Population

We filter the ACS data to private wage and salary workers in New York State with positive hours and wage income, removing the top and bottom 5% of earners to reduce the influence of extreme outliers. We then construct **annual hours worked** as our auxiliary variable. For this question we will consider this the population.

```{r}
ny_pop <- data |>
  filter(
    STATEFIP == 36,       # New York
    UHRSWORK > 0,
    INCWAGE  > 0,
    INCWAGE  < 999998,    # 999998/999999 are IPUMS N/A codes
    WKSWORK1 > 0,
    CLASSWKR == 2         # private wage/salary workers
  ) |>
  filter(
    INCWAGE < quantile(INCWAGE, 0.95),
    INCWAGE > quantile(INCWAGE, 0.05)
  ) |>
  mutate(annual_hours = UHRSWORK * WKSWORK1)

nrow(ny_pop)
```

------------------------------------------------------------------------

## Draw an SRS and Check the Key Assumption

We drew an SRS of size 1000 from the population. The ratio estimator requires a **proportional relationship** between $y$ and $x$.

```{r}

set.seed(123)
ny_sample <- ny_pop |> sample_n(1000)

B_hat = mean(ny_sample$INCWAGE)/mean(ny_sample$annual_hours)

# Scatterplot
ggplot(ny_sample, aes(x = annual_hours, y = INCWAGE)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = B_hat, col = "red", lwd = 1, linetype="dashed")+
  labs(
    title = "Wage Income vs. Annual Hours Worked (SRS, n = 1000)",
    x     = "Annual Hours Worked",
    y     = "Wage and Salary Income ($)"
  ) +
  theme_minimal()
```

> **Question 1:** Examine the scatterplot. Comment on whether or not you think the proportionality assumption is reasonable here. Comment on the strength of the correlation between hours worked and income.

## Efficiency of ratio estimation

The ratio estimator is more efficient than the sample mean estimator under SRS when:

$$R > \frac{1}{2} \cdot \frac{CV(x)}{CV(y)}$$

where $CV$ is the coefficient of variation and $R$ is the population correlation. Check whether this condition holds:

```{r}
cv_x      <- sd(ny_pop$annual_hours) / mean(ny_pop$annual_hours)
cv_y      <- sd(ny_pop$INCWAGE)      / mean(ny_pop$INCWAGE)
threshold <- 0.5 * (cv_x / cv_y)
R       <- cor(ny_pop$annual_hours, ny_pop$INCWAGE)
```

> **Question 2:** Does the efficiency condition hold? Interpret what this means in plain language.

------------------------------------------------------------------------

## Estimate the Ratio and Construct a Confidence Interval

We estimate $\hat{R}$ and construct a 95% CI under SRS using standard error formula:

$$\hat{V}(\hat{B}) = \left(1-\frac{n}{N}\right)\frac{s_e^2}{n\bar{x}_S^2}$$

where $s_e^2 = \frac{1}{n-1}\sum_{i\in S}(y_i - \hat{B}x_i)^2$

```{r}
n     <- _________
N     <- _________
fpc <-  __________
B_hat = _______________
xbar = _________________

se2 = __________________

Vhat = se2/(n*xbar^2)*fpc
SE = sqrt(Vhat)

ci_upr = B_hat + qt(0.975, df = 1000-1)*SE
ci_lwr = B_hat - qt(0.975, df = 1000-1)*SE

```

> **Question 3:** Interpret and give an estimate for $B$ in plain language and give a 95% confidence interval for $B$.

## Verification of results using the survey package

```{r}
# Set up SRS design
svy <- svydesign(
  ids     = ~1,          # no clustering
  fpc     = ~N,          # finite population correction
  data    = ny_sample |> mutate(N = nrow(ny_pop))
)

# Ratio estimate
ratio_svy <- svyratio(
  numerator   = ~INCWAGE,
  denominator = ~annual_hours,
  design      = svy
)

print(ratio_svy)
confint(ratio_svy, df = 999)
```

> **Question 4:** This is code to complete Q3 using the survey package. Verify that the estimate and the confidence interval you computed in Question 3 matches the one using this R package. (It should!)

## Ratio Estimator for the Population Total

Now we use $\hat{R}$ to estimate the **total wage income** of all private wage/salary workers in New York, and construct a CI for the total.

```{r}

t_x <- sum(ny_pop$annual_hours)  # known population total of annual hours
t_y  <- sum(ny_pop$INCWAGE)       # true total income (for verification)

# Point estimate
y_bar_hat_r <- ______________

# Variance of total:

Vhat = se2/n*(t_x/xbar)^2*fpc
SE = sqrt(Vhat)

y_bar_hat_r + qt(0.975, df = 1000-1)*SE
y_bar_hat_r - qt(0.975, df = 1000-1)*SE

```

> **Question 5:** What is the estimate for total wage income and a 95% CI? Does your 95% CI contain the true population total?

## Verification of results using the survey package

```{r}
# Set up SRS design
svy <- svydesign(
  ids     = ~1,          # no clustering
  fpc     = ~N,          # finite population correction
  data    = ny_sample |> mutate(N = nrow(ny_pop))
)

# Ratio estimate
ratio_svy <- svyratio(
  numerator   = ~INCWAGE,
  denominator = ~annual_hours,
  design      = svy
)

predict(ratio_svy, total = t_x)
```

> **Question 6:** This is code to complete Q6 using the survey package. Verify that the estimate and the standard error you computed in Question 6 matches the one using this R package. (It should!)

> **Question 7:** In this activity we used the full ACS dataset as the "known" population total $t_x$. In a real application where you only have a sample, where could you get $t_x$?

## Submission

Please submit this `.qmd` file rendered as a PDF, with all code chunks run and all questions answered. Make sure your file renders without errors before submitting.
